Higher-Order unification for free!

### 2024-05-03 paper/main.tex
proofreading

\begin{theorem}[Fidelity in $\wellb_\beta$] \Fullref{prop:simulation} and
\fullref{prop:fidelity} hold if $\;\wellb_\beta(\foUnifPb)$.
\label{th:fidelity-beta}
\end{theorem}
\begin{proof}[Proof sketch] Thanks to \cref{lem:w-enforcement-maybebeta}
  \Ue is as powerful as \Uo in \maybebeta, as well as in \wellb by
  \cref{prop:uniffowb}.
\end{proof}

@@

Each \linketa is kept in the link store \linkStore during execution
and is activated under some conditions.
Activation is implemented by extending the \elpiIn{progress1}
predicate defined in \cref{sec:execution}.

@@

When compiling a term $t$, we need to determine if any
subterm $s \in \subterm{t}$ that is of the form $\lambda x. r$,
where $x$ occurs in $r$, can be a $\eta$-expansion, i.e. if
there exists a substitution $\rho$ such that $\rho (\lambda x.r) \Eo s$.
The detection of lambda abstractions that can ``disappear''
is not as trivial as it may seems. Here a few examples:
%
\begin{center}
  \begin{tabular}{lll}
    %Term & Status & Evidence \\\hline
    $\lambda x. f \appsep (A \appsep x)$ & $\in\maybeeta$ & $\rho = \{~ A \mapsto \lambda x.x ~\}$ \\
    $\lambda x. f \appsep (A \appsep x) \appsep x$ & $\in\maybeeta$ & $\rho = \{~ A \mapsto \lambda x.a ~\}$\\
    $\lambda x. f \appsep x \appsep (A \appsep x)$ & $\not\in\maybeeta$ &\\
    $\lambda x. \lambda y. f \appsep (A \appsep x) \appsep (B \appsep y \appsep x)$ & $\in\maybeeta$ & $\rho = \{~ A \mapsto \lambda x.x~;~ B \mapsto \lambda y.\lambda x.y ~\}$
  \end{tabular}
\end{center}
\vspace{4pt}

% In the examples above, the first term is a \maybeeta since $A_x$ can
% reduce to $x$ by setting $A_x = \lambda x.x$, 
% the second one is not a \maybeeta since it exists no substitution
% for $A_x$ such that $A_x$ reduces to $x$ and $x$ is not free in the subterm $f \appsep x$.
\noindent
The first two examples are easy, and show how a unification variable can expose
or erase a variable in its scope, turning the resulting term into an $\eta$-expansion or not.\\
The third example shows that when a variable occurs outside the scope of a unification
variable, it cannot be erased and can hence prevent a term from being an $\eta$-expansion.\\
The last example shows the recursive nature of the check we need to implement.
The term starts with a spine of two lambdas, hence the whole term
is in \maybeeta iff the inner term $\lambda y.f\appsep (A \appsep x) \appsep (B \appsep y \appsep x)$
is in \maybeeta itself. If it is, it could $\eta$-contract to
$f \appsep (A \appsep x)$ making $\lambda x.f \appsep (A \appsep x)$ a potential
$\eta$-expansion.\\
We can now define more formally how \maybeeta terms are detected together with
its auxiliary functions:

@@

\begin{lemma}\label{lem:reduceto}
A $\beta$-normal term $s = \lambda x_1 \ldots x_n.t$
%, where $x, x_1\ldots x_n$ can occur in $t$, 
\reduceto{} $x$ only if one of the following three conditions holds:
\begin{enumerate}
  \item $n = 0$ and $t = x$;
  \item $t$ is the application of $x$ to a list
     of terms $l$ and each $l_i$ \reduceto{} $x_i$
     (e.g. $\lambda x_1 \ldots x_n.x \appsep x_1 \ldots x_n \Eo{} x$) ;
  \item $t$ is a unification variable with scope
    $W$, and for any $v \in \{ x, x_1 \ldots x_n \}$,
    there exists a $w \in W$, such that $w$ \reduceto{} $v$
    (if $n = 0$ this is equivalent to $x \in W$).
\end{enumerate}
\end{lemma}
\begin{proof}[Proof sketch]
Since our terms are in $\beta$-normal form there is only
one rule that can play a role (namely \ref{clause:eta1}),
hence if the term $s$ is not exactly $x$ (case 1)
it can only be an $\eta$-expansion of $x$, or a unification
variable that can be assigned to $x$, or a combination of both.
If $s$ begins with a lambda, then the lambda can only disappear by $\eta$
contraction. In that case, the term $t$ is under the spine of binders
$x_1\ldots x_n$ can either be $x$ applied to terms that can
\reduceto{} these variables (case 2), or a unification variable
that can be assigned to that application (case 3).
\end{proof}

@@

\noindent
Whenever \elpiIn{flam F} is detected to be in
\maybeeta it is compiled to \elpiIn{lam F1} and replaced by the fresh
variable \elpiIn{A}. This variable sees all the names free in
\elpiIn{lam F1} and is connected to \elpiIn{lam F1} via a \linketa.
Invariant~\fullref{inv:linklhs} holds for this link. Moreover:

@@

\begin{lemma}
  Let $\lambda x.t$ the \rhs of a \linketa, then $\wellb(t)$, enforcing \cref{inv:link-eta-right}.
  \label{lemma:unif-eta-aux}
\end{lemma}

\begin{proof}[Proof sketch]
  By construction, every ``problematic'' term in \Fo{} is replaced with a variable
  in the corresponding \Ho term. Therefore, $\wellb(t)$.
\end{proof}

@@

\begin{proof}[Proof sketch]
  Let $\sigma$ be the substitution, such that $\wellb(\sigma X)$ (since \Ue preserves \wellb).
  By \cref{inv:link-eta-right}, we have $\wellb(t)$.
  If $\sigma X = \lambda x.r$, then $r$ is unified with $t$ and both terms are in \wellb.
  Otherwise, $\lambda x. X x$ is unified with $\lambda x. t$, and $X x$ and $t$
  are both in \wellb.
\end{proof}

@@

\begin{proof}
  Given two \linketa{}s \linketaM{\Gamma_1}{X}{\lambda x.t} and
  \linketaM{\Gamma_2}{Y}{\lambda x.t'} 
  the unification is performed between $t$ and $t'$. Both terms are
  \wellb and therefore, unification is done between \wellb terms.
\end{proof}

@@

\begin{proof}[Proof sketch]
  Rules \cref{def:progressetaleft} and
  \cref{def:progressetadedup} remove one link from \linkStore, hence they
  cannot be applied indefinitely.
  Moreover each rule only relies on terminating operations such as \Ue,
  $\eta$-contraction, $\eta$-expansion, relocation (a recursive copy of a
  finite term).
  % The addition of rules for \elpiIn{progres1} complicates the function
  % \elpiIn{progress}. We can note, however, that they do not prevent the
  % termination of \elpiIn{progress}. 1) If a link is activated it is removed from
  % \linkStore and the recursive call to \elpiIn{progress} will have a smaller
  % list of links to recurse on. Moreover, link activation only runs terminating
  % instructions (such as unification). 2) If a link is deduplicated, the
  % termination of \elpiIn{progress} is still guaranteed since again we reduce
  % \linkStore and the instructions run by link deduplications are all
  % terminating. 3) If a link is neither activated nor deduplicated, i.e. it
  % remains suspended, then \linkStore remains unchanged like the substitution;
  % therefore, \elpiIn{if (L = L1, S1 = S2)} succeeds and \elpiIn{progress}
  % terminates.
\end{proof}

\begin{theorem}[Fidelity in $\wellb_{\beta \eta}$]
  % F x = z, F x y = z y
  Given a list of unification problems \foUnifPb, such that
  $\wellb_{\beta \eta}(\foUnifPb)$, if the 
  memory map is bijective, then
  the introduction 
  of \linketa guarantees \fullref{prop:fidelity}.
  \footnote{The bijective property of \mapStore is proved in \cref{sec:invariant1}}
  \label{lemma:fidelity-maybeeta}
\end{theorem}

\begin{proof}[Proof sketch] %\todo{to reformulate}
  % For any step $i$, we want to prove that $\fstep_i$ succeeds iff $\hstep_i$
  % succeeds. A $\hstep_i$ is made by a unification
  % $u = \sigma_{i-1}\hoUnifPb_{i_l} \Ue \sigma_{i-1}\hoUnifPb_{i_r} \mapsto \sigma_i'$ and a link
  % progression $p = \progress(\linkStore_{i-1},\sigma_i') \mapsto (\linkStore_i,\sigma_i)$.
  We want to prove that \fstep succeeds iff \hstep
  succeeds, where \hstep is made by a unification step $u$ and a link
  progression $p$.
  Without loss of generality we consider a
  \Fo unification problem of the form $s_1 \Uo s_2$
  where $s_1 \in \maybeeta$ and is compiled to a variable $X$ with the accessory link
  $\linketaM{\Gamma}{X}{\lambda x.t_1}$. 
  The unification $u$ always succeeds, since $X$ is a fresh variable, we 
  only have to consider the link progression $p$.
  Two cases should be analyzed:
  \begin{itemize}
    \item $s_2 \notin \maybeeta$,
          in this case $s_2$ is compile to $t_2$ and the unification step $u$ assigns $t_2$ to $X$
          and triggers \progressetaleft.
          If $s_1$ and $s_2$ unify thanks to \ref{clause:eta1} then
          progression $p$ unifies $\lambda x.t_1$ with the $\eta$-expansion of $t_2$ namly
          $\lambda x.t_2\appsep x$, otherwise they unify thanks to \ref{clause:lam-lam}
          and the link progress $p$ unifies $\lambda x.t_1$ with $t_2$.\\ Alternatively, if $s_1$ and $s_2$ fail
          to unify, it cannot be because of the $\lambda$ constructor in the head of $t_1$,
          since \ref{clause:lam-lam} or \ref{clause:eta1} try to unify under it, eventually finding two
          different subterms $s_1'$ and $s_2'$. By \cref{th:fidelity-beta},
          \Ue fails on terms $t_1'$ and $t_2'$ even if \progressetaleft
          $\eta$-expands $t_2$.

    \item $s_2 \in \maybeeta$, in this case, the unification problem becomes $X \Ue Y$ 
          with the extra link $\linketaM{}{Y}{\lambda x.t_2}$. After the unification of 
          $X$ and $Y$, \progressetadedup is fired and, once the $\lambda$ are crossed,
          $t_1$ and $t_2$ are unified. 
          This unification succeeds iff $s_1 \Uo s_2$ succeeds by \cref{th:fidelity-beta}.
  \end{itemize}
  % wlog we assume s1 is in \maybeeta, s2 is not (if both are thanks to dedup we are back in this situation).
  % we assume s1 is compiled to X and $\linketaM{\Gamma}{X}{t1}$.
  % In all cases $X \Ue t2$ succeeds since X is a variable and
  % hence assigns t2 to X.
  % Rule \progressetaleft unify with \Ue X (hence t2) with t1, with the only
  % additional care of eta expanding t2 if needed (5.9.1).
% 
  % Two cases should be analyzed: 1) $\fstep_i$ succeeds to unify 
  % $s1$ with $s2$. The interesting case is when fostep succeds by
  % finding a $\rho$ such that $\rho s1 \Eo \rho s2$ requires
  % rules~\ref{clause:eta1} and \ref{clause:eta2} at the top level,
  % and that is in exactly the cases in which T (from page 3) is
  % not a lam (see 5.9.1) compensating for the missing rules in \Ue.
  % 
  % 2) If $\fstep_i$ fails to unify $s1 \Uo s2$
  % it means that, after applying rules~\ref{clause:eta1} and \ref{clause:eta2}
  % and the one flam-flam we find different terms.
  % (5.9.1) only adds lambdas that are forcibly consumed by
  % lam-lam rule.
\end{proof}

@@

The example at the beginning of \cref{sec:eta}, once
$\sigma = \{~ A \mapsto f ~\}$, triggers \progressetaleft since the link
becomes \linketaM{}{f}{\lambda x.B_{x}} and the \lhs is a constant.
This rule runs $\lambda x.f \appsep x \Ue{} \lambda x.B_{x}$,
resulting in $\sigma = \{~ A \mapsto f ~;~ B_{x} \mapsto f ~\}$.
By the definition of \linketa decompilation,
decompilation generates $\rho = \{~ X \mapsto \lambda x. f \appsep x ~\}$, since:
$X$ is mapped to $B_x$ and $B_x$ is decompiled into $\lambda x.f \appsep x$.
% \todo{explain why decompile does so and which eta?}

@@

We want to allow for partial application of functions.
As a consequence
the same unification variable $X$ can be used multiple times
with different arities.
Then it is the case the
memory map \mapStore generated by the compiler is not
a bijection. For example:
\begin{comment}
%
% make test ONLY=7001 TEX=tex
\printAlll
  {{{\lambda x.\lambda y.(X\appsep y\appsep x),\lambda x.\lambda y.x},
    {\lambda x.(f\appsep (X\appsep x)\appsep x),Y}}}
  {{{A,\lambda x.\lambda y.x},
    {D,F}}}
  {{{X,E,1},
    {Y,F,0},
    {X,C,2}}}
  {{{\eta,,D,\lambda x.(f\appsep E_{x}\appsep x)},
    {\eta,,A,\lambda x.B_{x}},
    {\eta,x,B_{x},\lambda y.C_{y x}}}}
\end{comment}
% make test ONLY=7007 TEX=tex
\printAlll
  {{{\lambda x.(X\appsep x),a},
    {X,\lambda x.a}}}
  {{{A,a},
    {C,\lambda x.a}}}
  {{{X,C,0},
    {X,B,1}}}
  {{{\eta,,A,\lambda x.B_{x}}}}
% \todo{simplify example since no prog right}
in the unification problems \foUnifPb above, $X$ is used with arity $1$ in
$\foUnifPb_1$ and with arity $0$ in $\foUnifPb_2$. 
In order to preserve \fullref{inv:uvaarity} the compiler
did generate two entries in \mapStore for $X$, namely $B$ and $C$. 
However, there is no connection between these two variables
and any incompatible assignements to them would not be detected, breaking,
in turn \cref{prop:fidelity}. 
% This incompatibility will only be detected during the
decompilation phase.
To address this issue, we post-process the compiler's output
to ensure the following property:

\begin{proposition}[\mapStore is a bijection]
  After compilation, for each \Fo{} variable $X$ in \foUnifPb{} and for each \Ho{}
  variable $A$ in \hoUnifPb{} there is exactly one entry $X \mapsto A^n$
  (for any arity $n$).
\end{proposition}

@@

The intuition is that we $\eta$-expand the occurrence of the variable
with higher arity to match the lower arity. Since each \linketa can
add exactly one lambda, we need as many links as the difference between the
two arities.

@@

\begin{proof}[Proof sketch]
  By the definition of \mapdeduplication, any two occurrencies of the same \Fo{} 
  variable used at different arities are compiled to two
  variables $A^m$ and $C^n$ related by a chain of \linketa.
  % Attenzione a questo caso:
  %    A =𝜂 λx.(B x)   e    x |- B x =𝜂 λx.(C x)           (from a map dedup, where A e C, sono stessa variabile con diff arities)
  %  se (C x) è instanziato a (f (D x)) allora non c'è nessun progress:
  %  il rhs del secondo link diventa (λx.(f (D x))) che è un maybe-eta
%
  Without loss of generality, we pose $m < n$. If $A^m$ is assigned to a term
  $\lambda \vec{x}.t$ with $q$ being the length of $\vec{x}$. If $t$ is a rigid
  term, then the chain of links between $A^m$ and $C^n$ is completely triggered
  so that $C^n$ is assigned to $\lambda \vec{x}.t$ $\eta$-expanded $n-m$ times.
  Note that in this case fidelity is guaranteed, since term unified to $C^n$
  sohuld also unify with $A^m$. Otherwise, if $t$ is a variable, the chain of
  links is shrinked by $q$. Note that if this length is bigger then $n-m$ again
  the full chain collapse and $A^m$ and $C^n$ are agian tiny coupled. Otherwise,
  if $q < n-m$, it means that $C^n$ can always be assigned to any term starting
  with at least $q$ lambdas. In this second case fidelity is also guaranteed,
  due to the constraint of the number of lambda by which $C^n$ should start.
%
  % not starting with $\lambda$, then the entrire chain of link is
  % immediately waken up, assigning $C^n$ to $x_1\dots x_{n-m} \vdash t \appsep
  % x_1\dots x_{n-m}$, which is the $t$ $\eta$-expanded $n-m$ times. Hence, $A^m$
  % and $C^n$ are tiny coupled. If $A^m$ is assigned to a term of the form
  % $\lambda \vec{x}.t$, then if $t$ is rigid, the chain of links between $A^m$
  % and $C^n$ completely triggered, is shrinked by the length of $\vec{x}$. Note
  % that if $t$ is rigid, by the previous sentence, the remi
%
  % If $A^m$ is instantiated, 
%
  % Whenever the two variables are assigned to the non-unifiable terms,
  % the chain of \linketa is entirely waken up refining little by little
  % the intermediate variables between $A^m$ and $C^n$ leading eventually to a failure. 
  % On the other
  % hand, if the two variables still remain unifiable after their assignement, 
  % then the activation of 
  % the corresponding \linketa just refines the chain, by reducing eventually
  % its length.
\end{proof}
% \todo{Not sure I fullfy fixed the proof, but the old oen was not ok}

\noindent
Note that shortening the chain of links allows unification to fail if
$X_1$ and $X_2$ are unified with different terms.

@@

\noindent
Note that $\mapping{X}{B}{1}$ desappears from \mapStore
in favour of the auxiliary \linketa \linketaM{x}{E_{x}}{\lambda y.C_{x y}}.
The resolution of $\hoUnifPb_1$ assignes $a$ to $A$. This wakes up
$\linkStore_2$ by \progressetaleft, assigning <<$a \appsep x$>> to $B_x$.
The resolution of $\hoUnifPb_2$ instantiates $C$ to $\lambda x.a$, which, in
turn triggers $\linkStore_1$ by \progressetaleft. This link perfroms
$\lambda x.a \Ue \lambda x.(a \appsep x)$, which, as expected, fails.

@@

Even if we do not requrie \Ue to handle problems outside of \llambda it is
worth handling terms in \notllambda since, in practice, these terms often
re-enter \llambda at runtime, as observed in~\cite{Nadathur2001} for example,
or to accommodate for the introduction of heuristics.

>>>>

\begin{theorem}[Fidelity in $\wellb_\beta$] \Fullref{prop:simulation} and
\fullref{prop:fidelity} hold if $\;\wellb_\beta(\foUnifPb)$.
\label{th:fidelity-beta}
\end{theorem}
\begin{proof}[Proof sketch] Thanks to \cref{lem:w-enforcement-maybebeta}
  \Ue is as powerful as \Uo in $\wellb_\beta$, as well as in \wellb by
  \cref{thm:fidfow}.
\end{proof}

@@

Each \linketa is kept in the link store \linkStore during execution
and is activated under some conditions.
Activation is implemented
in \cref{sec:etaprogress}
by extending the \elpiIn{progress1}
predicate defined in \cref{sec:execution}.

@@

When compiling a term $t$, we need to determine if any
subterm $s \in \subterm{t}$ that is of the form $\lambda x. r$
can be an $\eta$-expansion, i.e. if
there exists a substitution $\rho$ such that $\rho (\lambda x.r) \Eo s$.
The detection of lambda abstractions that can ``disappear''
is not as trivial as it may seems. Here a few examples:
%
\begin{center}
  \begin{tabular}{lll}
    %Term & Status & Evidence \\\hline
    $\lambda x. f \appsep (A \appsep x)$ & $\in\maybeeta$ & $\rho = \{~ A \mapsto \lambda x.x ~\}$ \\
    $\lambda x. f \appsep (A \appsep x) \appsep x$ & $\in\maybeeta$ & $\rho = \{~ A \mapsto \lambda x.a ~\}$\\
    $\lambda x. f \appsep x \appsep (A \appsep x)$ & $\not\in\maybeeta$ &\\
    $\lambda x. \lambda y. f \appsep (A \appsep x) \appsep (B \appsep y \appsep x)$ & $\in\maybeeta$ & $\rho = \{~ A \mapsto \lambda x.x~;~ B \mapsto \lambda y.\lambda x.y ~\}$
  \end{tabular}
\end{center}
\vspace{4pt}

% In the examples above, the first term is a \maybeeta since $A_x$ can
% reduce to $x$ by setting $A_x = \lambda x.x$, 
% the second one is not a \maybeeta since it exists no substitution
% for $A_x$ such that $A_x$ reduces to $x$ and $x$ is not free in the subterm $f \appsep x$.
\noindent
The first two examples are easy, and show how a unification variable can expose
or erase a variable in its scope, turning the resulting term into an $\eta$-expansion or not.\\
The third example shows that when a variable occurs outside the scope of a unification
variable, it cannot be erased and can hence prevent a term from being an $\eta$-expansion.\\
The last example shows the recursive nature of the check we need to implement.
The term starts with a spine of two lambdas, hence the whole term
is in \maybeeta iff the inner term $\lambda y.f\appsep (A \appsep x) \appsep (B \appsep y \appsep x)$
is in \maybeeta itself. If it is, it could $\eta$-contract to
$f \appsep (A \appsep x)$ making $\lambda x.f \appsep (A \appsep x)$ a potential
$\eta$-expansion.\\
We can now detail how \maybeeta terms are detected.

@@

\begin{lemma}\label{lem:reduceto}
A $\beta$-normal term $s = \lambda x_1 \ldots x_n.t$
%, where $x, x_1\ldots x_n$ can occur in $t$, 
\reduceto{} $x$ only if one of the following three conditions holds:
\begin{enumerate}
  \item $n = 0$ and $t = x$;
  \item $t$ is the application of $x$ to a list
     of terms $l$ and each $l_i$ \reduceto{} $x_i$
     (e.g. $\lambda x_1 \ldots x_n.x \appsep x_1 \ldots x_n \Eo{} x$) ;
  \item $t$ is a unification variable with scope
    $W$, and for any $v \in \{ x, x_1 \ldots x_n \}$,
    there exists a $w \in W$, such that $w$ \reduceto{} $v$
    (if $n = 0$ this is equivalent to $x \in W$).
\end{enumerate}
\end{lemma}
\begin{proof}[Proof sketch]
Since our terms are in $\beta$-normal form there is only
one rule that can play a role (namely \ref{clause:eta1}),
hence if the term $s$ is not exactly $x$ (case 1)
it can only be an $\eta$-expansion of $x$, or a unification
variable that can be assigned to $x$, or a combination of both.
If $s$ begins with a lambda, then the lambda can only disappear by $\eta$
contraction. In that case, the term $t$ under the spine of binders
$x_1\ldots x_n$ can either be $x$ applied to terms that
\reduceto{} these variables (case 2), or a unification variable
that can be assigned to the application $x \appsep x_1 \ldots x_n$ (case 3).
\end{proof}

@@

\noindent
Whenever \elpiIn{flam F} is detected to be in
\maybeeta it is compiled to \elpiIn{lam F1} and replaced by the fresh
variable \elpiIn{A}. This variable sees all the names free in
\elpiIn{lam F1} and is connected to \elpiIn{lam F1} via a \linketa.
\Fullref{inv:linklhs} holds for this link. Moreover:

@@

\begin{lemma}
  Let $\lambda x.t$ be the \rhs of a \linketa, then $\wellb(t)$, enforcing \cref{inv:link-eta-right}.
  \label{lemma:unif-eta-aux}
\end{lemma}

\begin{proof}[Proof sketch]
  By \cref{lem:w-enforcement-maybeeta}.
\end{proof}

@@

\begin{proof}[Proof sketch]
  Let $\sigma$ be a substitution such that $\wellb(\sigma X)$ (since \Ue preserves \wellb).
  By \cref{inv:link-eta-right}, we have $\wellb(t)$.
  If $\sigma X = \lambda x.r$, then $r$ is unified with $t$ and both terms are in \wellb.
  Otherwise, $\lambda x. X x$ is unified with $\lambda x. t$, and $X x$ and $t$
  are both in \wellb.
\end{proof}

@@

\begin{proof}
  Given two \linketa{}s \linketaM{\Gamma_1}{X}{\lambda x.t} and
  \linketaM{\Gamma_2}{Y}{\lambda x.t'} 
  the unification is performed between $t$ and $t'$.
  By \cref{lem:w-enforcement-maybeeta} both terms are
  \wellb.
\end{proof}

@@

\begin{proof}[Proof sketch]
  Rules \ref{def:progressetaleft} and
  \ref{def:progressetadedup} remove one link from \linkStore, hence they
  cannot be applied indefinitely.
  Moreover each rule only relies on terminating operations such as \Ue,
  $\eta$-contraction, $\eta$-expansion, relocation (a recursive copy of a
  finite term).
  % The addition of rules for \elpiIn{progres1} complicates the function
  % \elpiIn{progress}. We can note, however, that they do not prevent the
  % termination of \elpiIn{progress}. 1) If a link is activated it is removed from
  % \linkStore and the recursive call to \elpiIn{progress} will have a smaller
  % list of links to recurse on. Moreover, link activation only runs terminating
  % instructions (such as unification). 2) If a link is deduplicated, the
  % termination of \elpiIn{progress} is still guaranteed since again we reduce
  % \linkStore and the instructions run by link deduplications are all
  % terminating. 3) If a link is neither activated nor deduplicated, i.e. it
  % remains suspended, then \linkStore remains unchanged like the substitution;
  % therefore, \elpiIn{if (L = L1, S1 = S2)} succeeds and \elpiIn{progress}
  % terminates.
\end{proof}

\begin{theorem}[Fidelity in $\wellb_{\beta \eta}$]
  % F x = z, F x y = z y
  Given a list of unification problems \foUnifPb such that
  $\wellb_{\beta \eta}(\foUnifPb)$, if the 
  memory map is bijective then
  the introduction 
  of \linketa guarantees \fullref{prop:fidelity}.
  \footnote{The bijective property of \mapStore is proved in \cref{sec:invariant1}}
  \label{lemma:fidelity-maybeeta}
\end{theorem}

\begin{proof}[Proof sketch] %\todo{to reformulate}
  % For any step $i$, we want to prove that $\fstep_i$ succeeds iff $\hstep_i$
  % succeeds. A $\hstep_i$ is made by a unification
  % $u = \sigma_{i-1}\hoUnifPb_{i_l} \Ue \sigma_{i-1}\hoUnifPb_{i_r} \mapsto \sigma_i'$ and a link
  % progression $p = \progress(\linkStore_{i-1},\sigma_i') \mapsto (\linkStore_i,\sigma_i)$.
  We want to prove that \fstep succeeds iff \hstep
  succeeds, where \hstep is made by a unification step $u$ and a link
  progression $p$.
  Without loss of generality we consider a
  \Fo{} unification problem of the form $s_1 \Uo s_2$
  where $s_1 \in \maybeeta$ and is compiled to a variable $X$ with the accessory link
  $\linketaM{\Gamma}{X}{\lambda x.t_1}$. 
  The unification $u$ always succeeds, since $X$ is a fresh variable, we 
  only have to consider the link progression $p$.
  Two cases should be analyzed:
  \begin{itemize}
    \item $s_2 \notin \maybeeta$,
          in this case $s_2$ is compiled to $t_2$ and the unification step $u$ assigns $t_2$ to $X$
          and triggers \progressetaleft.
          If $s_1$ and $s_2$ unify thanks to \ref{clause:eta1} then
          progression $p$ unifies $\lambda x.t_1$ with the $\eta$-expansion of $t_2$ namly
          $\lambda x.t_2\appsep x$, otherwise they unify thanks to rule \ref{clause:lam-lam}
          and the link progress $p$ unifies $\lambda x.t_1$ with $t_2$.\\ Alternatively, if $s_1$ and $s_2$ fail
          to unify, it cannot be because of the $\lambda$ constructor in the head of $t_1$,
          since rule \ref{clause:lam-lam} or rule \ref{clause:eta1} try to unify under it, eventually finding two
          different subterms $s_1'$ and $s_2'$. By \cref{th:fidelity-beta},
          \Ue fails on terms $t_1'$ and $t_2'$ even if \progressetaleft
          $\eta$-expands $t_2$.

    \item $s_2 \in \maybeeta$, in this case, the unification problem becomes $X \Ue Y$ 
          with the extra link $\linketaM{}{Y}{\lambda x.t_2}$. After the unification of 
          $X$ and $Y$, \progressetadedup is fired and, once the lambdas are crossed,
          $t_1$ and $t_2$ are unified. 
          This unification succeeds iff $s_1 \Uo s_2$ succeeds by \cref{th:fidelity-beta}.
  \end{itemize}
  % wlog we assume s1 is in \maybeeta, s2 is not (if both are thanks to dedup we are back in this situation).
  % we assume s1 is compiled to X and $\linketaM{\Gamma}{X}{t1}$.
  % In all cases $X \Ue t2$ succeeds since X is a variable and
  % hence assigns t2 to X.
  % Rule \progressetaleft unify with \Ue X (hence t2) with t1, with the only
  % additional care of eta expanding t2 if needed (5.9.1).
% 
  % Two cases should be analyzed: 1) $\fstep_i$ succeeds to unify 
  % $s1$ with $s2$. The interesting case is when fostep succeds by
  % finding a $\rho$ such that $\rho s1 \Eo \rho s2$ requires
  % rules~\ref{clause:eta1} and \ref{clause:eta2} at the top level,
  % and that is in exactly the cases in which T (from page 3) is
  % not a lam (see 5.9.1) compensating for the missing rules in \Ue.
  % 
  % 2) If $\fstep_i$ fails to unify $s1 \Uo s2$
  % it means that, after applying rules~\ref{clause:eta1} and \ref{clause:eta2}
  % and the one flam-flam we find different terms.
  % (5.9.1) only adds lambdas that are forcibly consumed by
  % lam-lam rule.
\end{proof}

@@

The example at the beginning of \cref{sec:eta}, once
$\sigma = \{~ A \mapsto f ~\}$, triggers \progressetaleft since the link
becomes \linketaM{}{f}{\lambda x.B_{x}} and the \lhs is a constant.
This rule runs $\lambda x.f \appsep x \Ue{} \lambda x.B_{x}$,
resulting in $\sigma = \{~ A \mapsto f ~;~ B_{x} \mapsto f ~\}$.
Since $X$ is mapped to $B_x$ and $B_x$ is decompiled into $\lambda x.f \appsep x$,
by the definition of \linketa decompilation
$\sigma$ is decompiled to $\rho = \{~ X \mapsto \lambda x. f \appsep x ~\}$,

% \todo{explain why decompile does so and which eta?}

@@

We want to allow for partial application of functions.
As a consequence
the same unification variable $X$ can be used multiple times
with different arities.
When it is the case the
memory map \mapStore generated by the compiler is not
a bijection. For example:
\begin{comment}
%
% make test ONLY=7001 TEX=tex
\printAlll
  {{{\lambda x.\lambda y.(X\appsep y\appsep x),\lambda x.\lambda y.x},
    {\lambda x.(f\appsep (X\appsep x)\appsep x),Y}}}
  {{{A,\lambda x.\lambda y.x},
    {D,F}}}
  {{{X,E,1},
    {Y,F,0},
    {X,C,2}}}
  {{{\eta,,D,\lambda x.(f\appsep E_{x}\appsep x)},
    {\eta,,A,\lambda x.B_{x}},
    {\eta,x,B_{x},\lambda y.C_{y x}}}}
\end{comment}
% make test ONLY=7007 TEX=tex
\printAlll
  {{{\lambda x.(X\appsep x),a},
    {X,\lambda x.a}}}
  {{{A,a},
    {C,\lambda x.a}}}
  {{{X,C,0},
    {X,B,1}}}
  {{{\eta,,A,\lambda x.B_{x}}}}
% \todo{simplify example since no prog right}
in the unification problems \foUnifPb above, $X$ is used with arity $1$ in
$\foUnifPb_1$ and with arity $0$ in $\foUnifPb_2$. 
In order to preserve \fullref{inv:uvaarity} the compiler
did generate two entries in \mapStore for $X$, namely $B$ and $C$. 
However, there is no connection between these two variables
and any incompatible assignements to them would not be detected,
in turn breaking \cref{prop:fidelity}. 
% This incompatibility will only be detected during the
% decompilation phase.
To address this issue, we post-process the compiler's output
to ensure the following property:

\begin{proposition}[\mapStore is a bijection]
  After compilation, for each \Fo{}-variable $X$ in \foUnifPb{} and for each
  \Ho{}-variable $A$ in \hoUnifPb{} there is exactly one entry $X \mapsto A^n$
  (for an arity $n$).
\end{proposition}

@@

The intuition is that we $\eta$-expand the occurrence of the variable
with lower arity to match the higher arity. Since each \linketa can
add exactly one lambda, we need as many links as the difference between the
two arities.

@@

\begin{proof}[Proof sketch]
  By the definition of \mapdeduplication, any two occurrencies of the same \Fo{} 
  variable used at different arities are compiled to two
  variables $A^m$ and $C^n$ related by a chain of \linketa.
  % Attenzione a questo caso:
  %    A =𝜂 λx.(B x)   e    x |- B x =𝜂 λx.(C x)           (from a map dedup, where A e C, sono stessa variabile con diff arities)
  %  se (C x) è instanziato a (f (D x)) allora non c'è nessun progress:
  %  il rhs del secondo link diventa (λx.(f (D x))) che è un maybe-eta
%
  Without loss of generality, we pose $m < n$. If $A^m$ is assigned to a term
  $\lambda \vec{x}.t$ with $q$ being the length of $\vec{x}$. If $t$ is a rigid
  term, then the chain of links between $A^m$ and $C^n$ is completely triggered
  so that $C^n$ is assigned to $\lambda \vec{x}.t$ $\eta$-expanded $n-m$ times.
  Note that in this case fidelity is guaranteed, since the term unified with $C^n$
  sohuld also unify with $A^m$. Otherwise, if $t$ is a variable, the chain of
  links is shrinked by $q$. Note that if this length is bigger then $n-m$
  the full chain collapse. % and $A^m$ and $C^n$ are agian tiny coupled. 
  Otherwise,
  if $q < n-m$, it means that $C^n$ can always be assigned to any term starting
  with at least $q$ lambdas. In this second case fidelity is also guaranteed,
  due to the constraint of the number of lambda by which $C^n$ should start.
%
  % not starting with $\lambda$, then the entrire chain of link is
  % immediately waken up, assigning $C^n$ to $x_1\dots x_{n-m} \vdash t \appsep
  % x_1\dots x_{n-m}$, which is the $t$ $\eta$-expanded $n-m$ times. Hence, $A^m$
  % and $C^n$ are tiny coupled. If $A^m$ is assigned to a term of the form
  % $\lambda \vec{x}.t$, then if $t$ is rigid, the chain of links between $A^m$
  % and $C^n$ completely triggered, is shrinked by the length of $\vec{x}$. Note
  % that if $t$ is rigid, by the previous sentence, the remi
%
  % If $A^m$ is instantiated, 
%
  % Whenever the two variables are assigned to the non-unifiable terms,
  % the chain of \linketa is entirely waken up refining little by little
  % the intermediate variables between $A^m$ and $C^n$ leading eventually to a failure. 
  % On the other
  % hand, if the two variables still remain unifiable after their assignement, 
  % then the activation of 
  % the corresponding \linketa just refines the chain, by reducing eventually
  % its length.
\end{proof}
\todo{this proof is still unreadable}
% \todo{Not sure I fullfy fixed the proof, but the old oen was not ok}

\noindent
Note that shortening the chain of links allows unification to fail if
$X_1$ and $X_2$ are unified with different terms.
\todo{what are these X1 and X2?}

@@

\noindent
Note that $\mapping{X}{C}{0}$ desappears from \mapStore
\todo{update this description to match the example} 
in favour of the auxiliary \linketa \linketaM{x}{E_{x}}{\lambda y.C_{x y}}.
The resolution of $\hoUnifPb_1$ assignes $a$ to $A$. This wakes up
$\linkStore_2$ by \progressetaleft, assigning <<$a \appsep x$>> to $B_x$.
The resolution of $\hoUnifPb_2$ instantiates $C$ to $\lambda x.a$, which, in
turn triggers $\linkStore_1$ by \progressetaleft. This link perfroms
$\lambda x.a \Ue \lambda x.(a \appsep x)$, which, as expected, fails.

@@

As observed in~\cite{Nadathur2001} it is worth 
handling terms in \notllambda since, in practice, these terms often
re-enter \llambda at runtime.
%for example,
%or to accommodate for the introduction of heuristics.
